\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-30 latex/article_vol2.Rnw
Rewritten part 2

Online tools such as OpenRefine (\url{http://openrefine.org/}) and
\proglang{R}-packages such as \pkg{plyr} \citep{plyr}, and
\pkg{data.table} \citep{data.table} have made data wrangling a lot
easier, but only a handful of packages such as \pkg{editrules}
\citep{editrules}, \pkg{validate} \citep{validate}, \pkg{DataCombine}
\citep{DataCombine}, \pkg{janitor} \citep{janitor}, and
\pkg{DataExplorer} \citep{DataExplorer} attempt to implement systematic,
reproducible data cleaning.  These packages use different approaches
for data cleaning: \pkg{editrules} and \pkg{validate} provide
frameworks for setting up and checking constraints on the variables,
while \pkg{DataCombine} and \pkg{janitor} both provide a few functions
for identifying problems (e.g, duplicates, dates coded as numbers,
etc.) in data.

While these tools attempt to alleviate the ubiquitous ad hoc approach
to data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a specific
field and the context of the available data. The \pkg{dataMaid}
package \citep{dataMaid} tries to address this by providing a framework that both allows for
extendable, systematic, reproducible data cleaning, and summarizing
findings for researchers from other fields such that they can act as
human experts when tracking down potential errors.

% \begin{itemize}
% \item Makes the point that data cleaning is usually done in a very ad hoc manner
% \item Maybe hints to the discussion later about poor documentation of data cleaning/quality assessment as the status quo for most analysts
% \item Gives a specification of what we mean by data errors/mistakes/whatever. I'm not completely sure this will be clear to everyone and we should probably mention it early on to not lose readers/users.
% \end{itemize}
% }

Data cleaning is a time consuming endeavor, as it inherently requires
human interaction since every dataset is different and the variables
in the dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{mySummaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

>>>>

A number of \proglang{R} packages made for such pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{plyr}
\citep{plyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data loading
with a particular emphasis on the challenges of getting neat
\code{data.frame}s from Microsoft Excel data files. \pkg{plyr},
\pkg{data.table} and \pkg{DataCombine} go a few steps further by
providing a wide array of extremely powerful tools for data wrangling,
including a number of particularly useful functions for merging and
working with very large datasets. When it comes to actual data
cleaning, however, the options are fewer.  \pkg{validate} (and the
similar packages \pkg{editrules} and \pkg{deducorrect} from the same
authors) does offer a few tools for identifying and auto-correcting
errors in a dataset, but, as the name suggests, the focus is on
internal validity rather than general data cleaning. In practice, this
means that quite elegant tools for e.g. linear restraints among the
variables in a dataset can be applied, but e.g. looking for
potentially miscoded missing values is not really feasible. The main
difference between these two challenges is the direction in which the
data is inspected: While linear constraints work row-wise with no
ambiguity, determining whether or not something is a miscoded missing
value often requires knowledge about the full variable (e.g. range or
data type), and thus it should be performed
column-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Moreover, no tools have been made for documenting the data cleaning
process.  As automatic cleaning options are also supplied in
\pkg{validate}, we find quite a dangerous cocktail where all power is
given to the the computer with no supervision, neither in real time
nor subsequently.

All in all, the large role of data cleaning in any data analysts
everyday endeavors is hardly matched in the amount of available
\proglang{R} software solutions. In particular, few packages attempt
to implement systematic, reproducible data cleaning. And while the
available tools attempt to alleviate the ubiquitous ad hoc approach to
data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a
specific field and the context of the available data. The
\pkg{dataMaid} package \citep{dataMaid} tries to address this by
providing a framework that both allows for extendable, systematic,
reproducible data cleaning, and summarizing findings for researchers
from other fields such that they can act as human experts when
tracking down potential errors.

One last package that should be mentioned in this context is
\pkg{DataExplorer}. While this package does not address data cleaning
issues per se, its general strategy is quite similar to that of
\pkg{dataMaid} and to the paradigms presented below. This package
provides a few simple, but practical, tools for exploratory data
analysis, including automated documentation. Therefore, we find
\pkg{DataExplorer} to be a good candidate for a next-step package
after data cleaning is finished.

But no matter how clever software tools we make, data cleaning remains
to be a time consuming endeavor, as it inherently requires human
interaction since every dataset is different and the variables in the
dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{summaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction()} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

### 2017-01-30 latex/article_vol2.tex
Rewritten part 1
Rewritten part 2

But no matter how clever software tools we make, data cleaning 
reamins to be a time consuming endeavor, as it inherently requires
human interaction since every dataset is different and the variables
in the dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

@@

As we saw with the \code{check()} function, the summary can be modified
by providing the relevant \code{XXXSummaries} arguments. Setting the
\code{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\code{numericSummaries = c("centralValue", "minMax")}:

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{mySummaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction()} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

@@

We will now build four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID()}] A new \code{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual()}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros()}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons()}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

@@

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID()}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros()}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColons()}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows: 
\begin{Schunk}
\begin{Sinput}
R> data("exampleData")
R> clean(exampleData,
+    preChecks = c("isKey", "isEmpty", "isID"),
+    allVisuals = "mosaicVisual",
+    characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
+    factorSummaries = c(defaultFactorSummaries(), "countZeros"),
+    labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
+    numericSummaries = c(defaultNumericSummaries(), "countZeros"),
+    integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
+    characterChecks = c(defaultCharacterChecks(), "identifyColons"),
+    factorChecks = c(defaultFactorChecks(), "identifyColons"),
+    labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}

>>>>

But no matter how clever software tools we make, data cleaning remains
to be a time consuming endeavor, as it inherently requires human
interaction since every dataset is different and the variables in the
dataset can only be understood in the proper context of their
origin. This often requires a collaborative effort between an expert
in the field and a statistician or data scientist, which may be why
the process of proper data cleaning is not always undertaken. In many
situations, these errors are discovered in the process of the data
analysis (e.g., a categorical variable with numeric labels for each
category may be wrongly classified as a quantitative variable or a
variable where all values have erroneously been coded to the same
value), but in other cases a human with knowledge about the data
context area is needed to identify possible mistakes in the data
(e.g., if there are 4 categories for a variable that should only have
3).

@@

As we saw with the \code{check()} function, the summary can be modified
by providing the relevant \code{XXXSummaries} arguments. Setting the
\code{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\code{numericSummaries=c("centralValue", "minMax")}:

@@

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Verbatim}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended
  for]))
\end{Verbatim}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{summaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction()} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

@@

We will now build four new functions and show both how they can be
used interactively and how they can be integrated with the \code{clean()}
function. These four new functions are:
\begin{description}
\item[\code{isID}] A new \code{checkFunction} intended for use in the precheck-stage. This function checks whether a variable consists exclusively of long ($> 10$ characters/digits) entries that are all of equal length, as this might be personal identification codes that we do not wish to print out in the data summary.
\item[\code{mosaicVisual}] A new \code{visualFunction} that produces mosaic plots. This function will be used in the \textit{visualize} step of \code{clean()}.
\item[\code{countZeros}] A new \code{summaryFunction} that counts the number of occurrences of the value \code{0} in a variable. This function will be used in the \textit{summarize} step of \code{clean()}.
\item[\code{identifyColons}] A new \code{checkFunction} that flags variables in which values have colons that appear before and after alphanumerical characters. This is practical for identifying autogenerated interaction effects. This function will be used in the \textit{check} step of \code{clean()}.
\end{description}
These functions are defined in turn below, and afterwards, an example of how they can be called from \code{clean()} is provided.

@@

\subsubsection{Calling the new summarize/visualize/check functions from clean()}
Now, we are ready to use these new functions in a \code{clean()}
call. The extended \pkg{dataMaid} output document should have the
following modifications, relative to the standard \pkg{dataMaid} output:
\begin{itemize}
\item We want to add the new pre-check function, \code{isID()}, to the already existing pre-checks.
\item We wish to change the plot type for all variables to the new mosaic plot.
\item We want the new summary function, \code{countZeros()}, to be added to the summaries performed on all variable types but \code{Date} and \code{logical}.
\item We want the new check function, \code{identifyColons()}, to be added to the checks performed on \code{character}, \code{factor} and \code{labelled} variables.
\end{itemize}
These options are specified as follows:

\begin{Schunk}
\begin{Sinput}
R> data("exampleData")
R> clean(exampleData,
+    preChecks = c("isKey", "isEmpty", "isID"),
+    allVisuals = "mosaicVisual",
+    characterSummaries = c(defaultCharacterSummaries(), "countZeros"),
+    factorSummaries = c(defaultFactorSummaries(), "countZeros"),
+    labelledSummaries = c(defaultLabelledSummaries(), "countZeros"),
+    numericSummaries = c(defaultNumericSummaries(), "countZeros"),
+    integerSummaries = c(defaultIntegerSummaries(), "countZeros"),
+    characterChecks = c(defaultCharacterChecks(), "identifyColons"),
+    factorChecks = c(defaultFactorChecks(), "identifyColons"),
+    labelledCheck = c(defaultLabelledChecks(), "identifyColons"))
\end{Sinput}
\end{Schunk}
